A quick analysis of Baltimore crime
========================================================

I'm going to do a very simple analysis of Baltimore crime to show off R. We'll use data downloaded from Baltimore City's awesome open data site. 

### Getting data

* Arrest data: https://data.baltimorecity.gov/Crime/BPD-Arrests/3i3v-ibrt 
* CCTV data: https://data.baltimorecity.gov/Crime/CCTV-Locations/hdyb-27ak 

Let's load the data:
```{r}
arrest_tab=read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)
cctv_tab=read.csv("CCTV_Locations.csv", stringsAsFactors=FALSE)

# these columns are mislabeled, so fix them
tmp=arrest_tab$sex
arrest_tab$sex=arrest_tab$race
arrest_tab$race=tmp
```

### Exploring data

```{r}
# dimension of table (data.frame)
dim(arrest_tab)

# what are the columns
names(arrest_tab)

# what is the average arrest age?
mean(arrest_tab$age)

# the range of arrest ages
range(arrest_tab$age)

# how many arrests per sex
table(arrest_tab$sex)

# what are the most common offenses
head(sort(table(arrest_tab$incidentOffense),decreasing=TRUE))

# range of arrests after removing those w/ age==0
range(arrest_tab$age[arrest_tab$age>0])
```

Let's see a table of arrests by sex and race
```{r}
table(sex=arrest_tab$sex,race=arrest_tab$race)
```

A histogram of age

```{r}
hist(arrest_tab$age,nc=100)
with(arrest_tab,hist(age[sex=="M"],nc=100)) # males only
with(arrest_tab,hist(age[sex=="F"],nc=100)) # females only

```

### Are males and females arrested at different ages on average?

Let's take a look at how age depends on sex. Let's plot age as a function of sex first (notice how we indicate that sex is a `factor`). 

```{r}
plot(arrest_tab$age~factor(arrest_tab$sex))
```

One of the neat things about R is that statistical model building and testing is built-in. The model we use is $y_i=\beta_0+\beta_1 x_i$ where $y_i$ is age of sample (example) $i$ and $x_i$ is an indicator variable $x_i \in \{0,1\}$ with $x_i=1$ if the $i$-th record (example) is male. You can check that $\beta_1$ is the difference in mean age between females and males.
We use the formula syntax to build a linear regression model. 

```{r}
# let's ignore those records with missing sex
fit=lm(age~factor(sex),data=arrest_tab,subset=arrest_tab$sex %in% c("M","F"))
summary(fit)
```

We see that $\beta_1 \approx -0.2$ meaning that the arrest age for males is about 2.5 months younger. So there is very little difference in the average age (which is what the linear model is testing) but we see that the probability of observing this difference from a sample of this size **when there is no difference in average age** is small $p \approx 0.01$. Since we have a very large number of examples, or records, this testing framework will declare very small differences as *statistically significant*. We'll return to this theme later in class.


### Geographic distribution of arrests.

First we need to extract latitude and longitude from location, we'll use some string functions to do this

```{r}
tmp=gsub("\\)","",gsub("\\(","",arrest_tab$Location))
tmp=strsplit(tmp,split=",")
arrest_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
arrest_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))
```

Now let's plot

```{r}
plot(arrest_tab$lon, arrest_tab$lat, xlab="Longitude", ylab="Latitude", main="Arrests in Baltimore")
```

We can also use density estimates to make this nicer:

```{r}
smoothScatter(arrest_tab$lat, arrest_tab$lon, xlab="Latitude", ylab="Longitude", main="Arrests in Baltimore")
```

Let's make this fancier using the `ggplot2` graphics systems and the `maps` package containing map data.

```{r}
library(maps)
library(ggplot2)

balto_map = subset(map_data("county", region="maryland"),subregion=="baltimore city")
plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=arrest_tab,aes(x=lon,y=lat),color="blue",alpha=.1)
print(plt)
```

Now let's add CCTV cameras.

```{r}
tmp=gsub("\\)","",gsub("\\(","",cctv_tab$Location))
tmp=strsplit(tmp,split=",")
cctv_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
cctv_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))

plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=arrest_tab,aes(x=lon,y=lat),color="blue",alpha=.1)
plt=plt+geom_point(data=cctv_tab,aes(x=lon,y=lat),color="red")
print(plt)
```

### A challenge

Is there any relationship between the number of CCTV cameras and the number of arrests? Divide the city into a grid and plot the number of CCTV cameras vs. the number of arrests.

```{r}
latRange=range(arrest_tab$lat,na.rm=TRUE)
lonRange=range(arrest_tab$lon,na.rm=TRUE)
latGrid=seq(min(latRange),max(latRange),len=50)
lonGrid=seq(min(lonRange),max(lonRange),len=50)
latFac=as.numeric(cut(arrest_tab$lat,breaks=latGrid))
lonFac=as.numeric(cut(arrest_tab$lon,breaks=lonGrid))

gridFac=(latFac-1)*length(latGrid) + (lonFac-1)

latFac=as.numeric(cut(cctv_tab$lat,breaks=latGrid))
lonFac=as.numeric(cut(cctv_tab$lon,breaks=lonGrid))
cctvGridFac=(latFac-1)*length(latGrid)+(lonFac-1)

arrestTab=table(gridFac)
cctvTab=table(cctvGridFac)
m=match(names(cctvTab),names(arrestTab))
plot(arrestTab[m]~factor(cctvTab))
```

### Extra analyses

As part of HW1 you will add to this analysis. Please use the following template:

#### Your name(s) here
Eric Krokos

What question are you asking?:
What does the distribution of arrests look like for Asians(Blue), Blacks(Red), Whites(Green), and Unknown (Orange)?

What is the code you use to answer it?:

```{r EKrokos}
asian = arrest_tab[arrest_tab$race == "A",]
black = arrest_tab[arrest_tab$race == "B", ]
white = arrest_tab[arrest_tab$race == "W", ]
unknown = arrest_tab[arrest_tab$race == "U", ]
library(maps)
library(ggplot2)
balto_map = subset(map_data("county", region="maryland"),subregion=="baltimore city")
plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=asian,aes(x=lon,y=lat),color="blue",alpha=.1)
plt=plt+geom_point(data=black,aes(x=lon,y=lat),color="red",alpha=.1)
plt=plt+geom_point(data=white,aes(x=lon,y=lat),color="green",alpha=.1)
plt=plt+geom_point(data=unknown,aes(x=lon,y=lat),color="orange",alpha=.1)
print(plt)
```

What did you observe?
=======
The red (Black) dominate the map, but it's interesting that most crimes commited by Whites are done on the southern most part of the city, with Black crime occuring primarily in the upper part. It's difficult to see Asian commited crime. Given that most of the orange dots are in the southern part of the city, I would guess that most of the unknown races are from White people.

#### Andy Garron, Mohit Iyyer, Peter Enns, Fan Du

What question are you asking?: In which months do arrests occur the most? On what day of the week do arrests occur the most?

What is the code you use to answer it?:

```{r ClouchPotatoes}
arrestDates = as.Date(arrest_tab$arrestDate,"%m/%d/%Y")
dwka <- format(arrestDates , "%a")
#dwka
dwkn <- as.numeric(format(arrestDates,"%w")) # numeric version
hist( dwkn , main = "Arrests by Day of Week", xlab = "Day of Week", ylab = "Arrest Frequency", breaks= -.5+0:7, labels= unique(dwka[order(dwkn)]))

dwkm <- format(arrestDates, "%b") # get month labels
dwkmn <- as.numeric(format(arrestDates,"%m")) # get numeric data for histogram
hist(dwkmn, main = "Arrests by Month",xlab = "Month of Year",ylab = "Arrest Frequency",breaks = 0:12 ,labels= unique(dwkm[order(dwkmn)]))
```

At the month level, we observe that there is a decline in arrest rates during the winter months and that arrests peak in March and August.

At the day level, we observe that most arrests occur in the middle of the week and that significantly fewer arrests occur on the weekends.

#### Patricia Sazama

Is there a relationship between the age of the arrested individual and the time they are arrested?

```{r psazama}
#change the time format from hh:mm to hh.mm so it can be converted to a numeric value
tmp=gsub(":",".",arrest_tab$arrestTime)

#convert the arrestTime to numeric
arrestTime=as.numeric(tmp)

#group ages into 5 buckets
ages = arrest_tab$age
ages = cut(ages, 5)

#plot the age buckets as a factor of the numeric arrest times
plot(arrestTime~factor(ages))
```
From these plots we can see slight differences in the average times of day members of these age groups are arrested.  We can observe that as age increases, generally the time of arrest is earlier in the day though this is only a very slight difference.  For all groups the average time of arrest is very close to 15:00.


#### Steven Burgart

What question are you asking?: Is there a relationship between the time a person is arrested and their sex?

What is the code you use to answer it?:

```{r sburgart}

# Make a copy of arrest_tab since we need to modify
arrest_tab_skb <- arrest_tab

# Filter out arrests with unknown sex
arrest_tab_skb <- arrest_tab_skb[arrest_tab_skb$sex=="M" | arrest_tab_skb$sex=="F",]

# Convert time to just hours as numeric
arrest_tab_skb["arrestTime"] <- lapply(arrest_tab_skb["arrestTime"], function(x) as.numeric(strftime(strptime(x, format="%H:%M"),"%H")))

# Plot male arrests by hour
hist(arrest_tab_skb[arrest_tab_skb$sex=='M',]$arrestTime, breaks=24, main="Hour of Arrest Frequency for Males", xlab="Hour of the Day", col="lightblue", xlim=range(0:23), xaxt='n');
axis(side=1, at=seq(0,23), labels=seq(0,23))

# Plot female arrests by hour
hist(arrest_tab_skb[arrest_tab_skb$sex=='F',]$arrestTime, breaks=24, main="Hour of Arrest Frequency for Females", xlab="Hour of the Day", col="pink", xlim=range(0:23), xaxt='n');
axis(side=1, at=seq(0,23), labels=seq(0,23))

# Plot relationship between hour of arrest and sex
plot(arrest_tab_skb$arrestTime ~ factor(arrest_tab_skb$sex), main="Relationship Between Hour of Arrest and Sex", xlab="Sex", ylab="Hour of Arrest")
```


What did you observe?: From the two histograms it isn't readily apparent that there is any relationship between the time a person is arrested and their sex. However, from the third plot we can see that males tend to be arrested slightly later in the day when compared to females.


#### Rob Argue

What question are you asking?:

Do violent crimes occur more in certain areas?


What is the code you use to answer it?:

```{r RobArgue}

# The FBI classifies the following as violent crime:
#  - murder and nonnegligent manslaughter
#  - forcible rape
#  - robbery
#  - aggravated assault
# as per
# http://www.fbi.gov/about-us/cjis/ucr/crime-in-the-u.s/2011/crime-in-the-u.s.-2011/violent-crime/violent-crime

violent = c("1A-Murder",
            "2A-Rape (Force)",
            "2B-Rape (Attempt)",
            "3AF-Robb Hwy-Firearm",
            "3AJF-Robb Carjack-Firearm",
            "3AJK-Robb Carjack-Knife",
            "3AK-Robb Hwy-Knife",
            "3AO-Robb Hwy-Other Wpn",
            "3CF-Robb Comm-Firearm",
            "3CK-Robb Comm-Knife",
            "3CO-Robb Comm-Other Wpn",
            "3EF-Robb Gas Station-Firearm",
            "3EK-Robb Gas Station-Knife",
            "3GF-Robb Conv Store-Firearm",
            "3GK-Robb Conv Store-Knife",
            "3GO-Robb Conv Store-Other Wpn",
            "3JF-Robb Residence-Firearm",
            "3JK-Robb Residence-Knife",
            "3JO-Robb Residence-Other Wpn",
            "3LF-Robb Bank-Firearm",
            "3FO-RObb Bank-Other Wpn",
            "3NF-Robb Misc-Firearm",
            "3NK-Robb Misc-Knife",
            "3NO-Robb Misc-Other Wpn",
            "4A-Agg. Asslt.- Gun",
            "4B-Agg. Asslt.- Cut",
            "4C-Agg. Asslt.- Oth.",
            "4D-Agg. Asslt.- Hand")

plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=arrest_tab,aes(x=lon,y=lat),color="blue",alpha=.1)
plt=plt+geom_point(data=arrest_tab[arrest_tab$incidentOffense %in% violent,],aes(x=lon,y=lat),color="red",alpha=.2)
print(plt)
```

What did you observe?

It appears that there is a heavier concentration of violent crimes in the inner city, particularly in a few hotspots (of which I am unaware the significance of). There also appear to be a couple of specific outlying locations where an unusually large number of violent crimes occur.
#### Hao Zhou(zhhoper), Fang Cheng(Javran)

* What question are you asking?

    Is there any difference between minors and adults
    in terms of time and location?
    (we suppose people under 21 are minors)

* What is the code you use to answer it?
    
    We split data into two parts
    (i.e.  minors (age < 21) and adults (age >= 21)),
    plot the distribution of arrest time 
    and draw scatters according to the location
    for both minors and adults.

```{r zhhoper_and_Javran}

# copy data frame,
#   filter out invalid data (age = 0)
arrestTmp <- subset(arrest_tab, arrest_tab$age !=0)

# extract geo info
tmp <- gsub("\\).*", "", gsub(".*\\(", "", arrestTmp$Location))
tmp = strsplit(tmp, split = ",")
arrestTmp$lat <- as.numeric(sapply(tmp, function(x) x[2]))
arrestTmp$lon <- as.numeric(sapply(tmp, function(x) x[1]))

# suppress warnings
arrestTmp <- subset(arrestTmp, !is.na(lon) & !is.na(lat))

# tag data with "minors" and "adults" according to their ages
arrestTmp$biAge <- sapply(arrestTmp$age, function(x) {
  if (x < 21) {return("minors")}
        else  {return("adults")}})

# new col: time -> time in Hour
arrestTmp$arrestTimeH <- as.integer(
  gsub(":[0-9][0-9]","",
  arrestTmp$arrestTime))

# plot minors' distribution
minors_tab <- subset(arrestTmp, arrestTmp$biAge == "minors")
minorsArray <- minors_tab$arrestTimeH
minorsTable <- as.integer(table(factor(minorsArray)));
barplot(minorsTable, space = 0, width = 1,
        xlab = "Time", ylab="Number of arrset",
        main="Number of arrest in each hour (minors)", axes = T)
axis(side = 1, at = seq(0, 23), labels = seq(0, 23))

# plot adults' distribution
adults_tab <- subset(arrestTmp, arrestTmp$biAge == "adults")
adultsArray <- adults_tab$arrestTimeH
adultsTable <- as.integer(table(factor((adultsArray))));
barplot(adultsTable, space = 0,width = 1,
        xlab = "Time", ylab="Number of arrset",
        main="Number of arrest in each hour (adults)",axes = T)
axis(side = 1, at = seq(0, 23), labels = seq(0, 23))

plot(arrestTmp$arrestTimeH~factor(arrestTmp$biAge),
    main = "Relationship Between time of Arrest and minor or not",
    xlab = "age", ylab = "Hour of Arrest")

# visualize arrest location for minors & adults
smoothScatter(minors_tab$lat, minors_tab$lon,
              xlab = "Latitude", ylab = "Longitude",
              main = "Arrests in Baltimore (minors)")
smoothScatter(adults_tab$lat, adults_tab$lon,
              xlab = "Latitude", ylab = "Longitude",
              main = "Arrests in Baltimore (adults)")

# cleanup
rm(arrestTmp,tmp)
rm(adults_tab,adultsTable,adultsArray)
rm(minors_tab,minorsTable,minorsArray)


```

* What did you observe?

    * Through the first two figures, we observe that
    minors tend to commit less during 1 - 6 a.m., but more during 2 - 8 p.m.

    * There is little difference between the time distribution of minors and adults.

    * From the last two scatters we can tell that the density is slightly different in the northwest part.
    More interesting conclusions could be drawn if we can compare this result with the local map.

#### Hui Miao

What question are you asking?: What is the most common incident offense, and how does it related to sex, race and district?

What is the code you use to answer it?:

```{r imoldcat}
old_par <- par(no.readonly=TRUE)

# first let's look at the bar plot of the distribution of incident offense types
# as there are many # of inccident offsense types, I use a filter to get rid of 
# 'unknown offense' that we don't care about, and the ones with less than 200 cases
# in order to make the plot clear to understand
library(sqldf)
a = sqldf("select incidentOffense, count(*) as cnt from arrest_tab where incidentOffense != 'Unknown Offense' group by incidentOffense having cnt >= 200 order by cnt")
par(las=2, mar=c(5,7,4,2))
barplot(a$cnt, horiz=TRUE, cex.names=0.7, names.arg=a$incidentOffense)

# as you can see, the most common incident offense is 'Narcotics' (the 1st and 3rd row)
# next, let's analyze the correlations between 'Narcotics' inccident type and sex, race and district
# this is a correlation analysis invovling in multiple categorical variables
# I use a mosaic plot instead of the correlogram plot to show their correlations in one plot

library(vcd)
filtered <- subset(arrest_tab, (race == "B" | race == "W") & sex != "" & district != "" & (incidentOffense == "87-Narcotics" | incidentOffense == "87O-Narcotics (Outside)"))
filtered$district <- gsub("NORTHERN","N", gsub("NORTHEASTERN","NE", gsub("NORTHWESTERN","NW",filtered$district)))
filtered$district <- gsub("SOUTHERN","S", gsub("SOUTHEASTERN","SE", gsub("SOUTHWESTERN","SW",filtered$district)))
filtered$district <- gsub("WESTERN","W", gsub("EASTERN","E", gsub("CENTRAL","C",filtered$district)))
mosaic(~ district + sex + race + incidentOffense , data=filtered, shade=TRUE, legend=TRUE)
par(old_par)
```

What did you observe?: The first bar plot shows that the 'Narcotics' is the most common inccident offense in this city. Interestingly 'Narcotics (Outside)' is another top3 inccident offense type.  

In the second plot, I use mosaic plot to visualize the correlations among multiple categorical variables (race (b, w), sex (m, f), district (w, sw, se, s, nw, ne, n, e, c)). Rectangle sizes reflect the frequentices of a group, the colors shows the pearson residuals (blue means above expected value, while red means it's below expected value of fitted model).

Here're some interesting findings:

1. From previous map plots with race info, we know that in the southern regions, white cases and black cases are similar. It is the same case for Narcotics. The south regions (SE, S), the numbers are similar (we can see from the heighs of the rectangle). In other regions, black dominants white. 

2. Where you can find narcotics cases? And where does outside Narcotics case happen more often than indoor? North (N), Southerne East (SE) and South (S) regions have the fewest instances of Narcotics cases. In the north eastern (NE) and south western (SW) regions, the Narcotics outside cases are above expected value, while Eastern (E), Northwestern (NW) and Western (W) regions outside cases are below expected value. 

3. Female's narcotics active region is quite different from male. In the central district (C), both female black and white are above the expected values for both inside and outside narcotics cases, while male cases regardless races are below expected value.


#### Ben Klimkowski and Jon Fetter-Degges
What question are you asking?:

What neighborhoods have the highest number of arrests? Of narcotics arrests? What is the nature of the crime in the worst neighborhoods?

What is the code you use to answer it?:
```{r Klimkowski_and_Fetter_Degges}
# We begin by counting the total number of arrest records per neighborhood, throwing
# out those where the neighborhood is empty.
total <- table(arrest_tab$neighborhood[arrest_tab$neighborhood != ""])
# Now we do it again, using only narcotics arrests.
narc.arrests <- arrest_tab$neighborhood[grep("narcotics", 
                                     arrest_tab$incidentOffense, ignore.case=TRUE)]
narc <- table(narc.arrests[narc.arrests != ""])
# We want to put these vectors into the same data frame, so let's throw away
# neighborhoods with no narcotics arrests.
total <- total[names(total) %in% names(narc)]
df <- data.frame(total)
df$narc <- narc
# Now we can sort the frame by total number of arrests.
df <- df[order(df$total, df$narc), ]
df$nonnarc <- df$total - df$narc
# We'll get rid of extra variables; everything we need is in the frame.
rm(total, narc, narc.arrests)
# And generate a plot of neighborhoods with the most arrests, along with how many of
# them are for narcotics.
# We looked at Hui Miao's code to find out how to save and restore graph parameters.
old_par <- par(no.readonly = TRUE)
par(las=2, mar=c(4, 9, 3, 2)+0.1)
barplot(t(as.matrix(df[df$total>600, 2:3])), horiz=TRUE, col=c("orange", "blue"), cex.names=0.7, main="Number of Arrests per Neighborhood--")
legend(2000, 12, legend=c("narcotics", "non-narcotics"), fill=c("orange", "blue"), cex=0.7)
par(old_par)

#Using the code above it, we will now create graphics for the 5 neighborhoods with the most crime:
##Downtown, Sandtown-Winchester, Central Park Heights, Broadway East, Belair-Edison

library(plotrix)
bmoreHoodAnalyze<-function(arg1){
  #This function cleans the Baltimore dataset by combining similar crimes and
  #and graphically depicts the most frequently crimes in an area
  a<-table(arrest_tab$incidentOffense[arrest_tab$neighborhood==arg1])
  df<-data.frame(a)
  
  #Combine sums
  ##Narcotics
  narcoSum<-sum(df$Freq[df$Var1=="87O-Narcotics (Outside)"| df$Var1=="87-Narcotics"])
  df$Freq[df$Var1 == '87-Narcotics']<-narcoSum
  df<-df[!df$Var1 == '87O-Narcotics (Outside)',]
  ##Unlabeled crimes
  unkSum<-sum(df$Freq[df$Var1=="Unknown Offense"| df$Var1=="79-Other"])
  df$Freq[df$Var1 == 'Unknown Offense']<-unkSum
  df<-df[!df$Var1 == '79-Other',]
  df<-df[!df$Var1 == 'Unknown Offense',]

  #Concatenate the remainder of the crimes
  df <-df[order(-df$Freq),]
  l <-as.vector(df[,1])[0:5]
  freq <-as.vector(df[,2])[0:5]
  freq<-c(freq,sum(as.vector(df[,2])[6:dim(df)[1]])+unkSum)
  l<-c(l,"Other")
  title<-paste0("Distribution of Type of Crime in ", arg1)
  pie3D(freq,labels=l, explode=.2,theta=pi/2.5,labelcex=.7, main=title)
}


bmoreHoodAnalyze('Downtown')
bmoreHoodAnalyze('Sandtown-Winchester')
bmoreHoodAnalyze('Central Park Heights')
bmoreHoodAnalyze('Broadway East')
bmoreHoodAnalyze('Belair-Edison')


```
What did you observe?:

It is clear from this analysis that narcotics related crimes are the most frequent ones across many neighborhoods in Baltimore.

Unfortunately, the Baltimore database left many entries as "79-Other" or unknown, so specific crimes are hard to correlate.  However, with some cleaning of the data-set, a clear pattern emerged in these problem spots.  After narcotics, the most frequent crime to be encountered would be assault or search/seizure related, followed by towed vehicles.